---
title: "Practical Machine Learning : Peer Graded Assingment"
author: "kamal pradhan"
date: "29 October 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to predict the manner in which they did the exercise.

##Data

The training data for this project are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)

The test data are available here:

[https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)


The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

###Setting up the Environment
This portion covers the loading of all libraries required for the project.
```{r p_setup, warning=FALSE, results="hide", message=FALSE}
#loading libraries
library(ggplot2)
library(dplyr)
library(caret)
library(corrplot)
library(reshape2)
library(randomForest)
```


First we will download the ***training*** and ***testing*** dataset from the links provided above. 
```{r g_data}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if(!file.exists("pml-training.csv") || !file.exists("pml-testing.csv")){
  download.file(train_url,method = "curl",destfile = "pml-training.csv")
  download.file(test_url,method = "curl",destfile = "pml-testing.csv")
  downloadDate<- date()
  
  train_data<-read.csv("pml-training.csv",sep = ",",header = TRUE,na.strings = c("",NA))
  test_data<-read.csv("pml-testing.csv",sep = ",",header = TRUE,na.strings = c("",NA))
 
}else{
  train_data<-read.csv("pml-training.csv",sep = ",",header = TRUE,na.strings = c("",NA))
  test_data<-read.csv("pml-testing.csv",sep = ",",header = TRUE,na.strings = c("",NA))
}

```

###Data Preparation

As both the test and train data contains maximum rows empty columns.So we remove the columns where 50% of the rows are empty. we also remove variables such as username, timestamp e.t.c. which are not necessary.

```{r d_prep}
#Marking columns for removal
rm_cols<-colnames(train_data[,colSums(is.na(train_data)) > nrow(train_data)*0.50])

#Removing from training set
train_data<-train_data[,!names(train_data) %in% rm_cols]

#Removing from test set
test_data<- test_data[,!names(test_data) %in% rm_cols]

#Marking Unnecessary columns
un_cols<- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2","cvtd_timestamp", "new_window", "num_window")

#Removing Unnecessary columns from training set
train_data<-train_data[,!names(train_data) %in% un_cols]

#Removing Unnecessary columns from test set
test_data<- test_data[,!names(test_data) %in% un_cols]
```

###Data Exploration and Analysis

Here we will explore the nature of the dataset and try to extract all the meaningful insights from it.

```{r d_explor}
#dimension of training and testing set
dim(train_data)
dim(test_data)

#class variable factors
table(train_data$classe)

#correlation between variables(top 20)
corrs<-cor(train_data[,-53])
corrs[corrs==1]<-NA
corrs[abs(corrs)<0.5]<-NA
corrs<-na.omit(melt(corrs))
corrs[order(-abs(corrs$value)),][1:20,]
```

These are some highly correlated variables.

###Splittinng the Data before building the models
For the comutation of out of sample error we divide the training data set in 7:3 ratio for training and validation.

```{r d_split}
set.seed(3421) 

#dividing data for training and validation
temp_train <- createDataPartition(train_data$classe, p = 0.7, list = FALSE)
train <- train_data[temp_train, ]
valid <- train_data[-temp_train, ]
```

###Building and Evaluation of Training Models
For the project i will use two models
 
 * Decision tree model
 * Random forest model
 
Here we use, K=5, when doing k-fold cross validation since we are using non-linear models.
```{r M_build, warning=FALSE, results="hide", message=FALSE}
control <- trainControl(method = "cv", number = 5)

#Decision Tree Model
dtree_model <- train(
  classe ~ ., 
  data=train,
  trControl=control,
  method='rpart'
)

#Random Forest Model
rforest_model <- train(
  classe ~ ., 
  data=train,
  trControl=control,
  method='rf',
  ntree=100
)


```

###Model Accuracy Comparision
```{r r_comp}
#predicting using validation dataset and getting accuracy
dtree_predict<-predict(dtree_model,valid)
rforest_predict<-predict(rforest_model,valid)

#confusion Matrix for decision tree
d_mat<-confusionMatrix(dtree_predict,valid$classe)
print(d_mat)

#confusion Matrix for Random Forest
rf_mat<-confusionMatrix(rforest_predict,valid$classe)
print(rf_mat)

#Accuracy of decision tree model
d_mat$overall[1]

#Accuracy of Random forest model
rf_mat$overall[1]

```

###Predicting the Testing set
Since Random Forest Model is Having more accuracy than the decision tree Model. We will use Random Forest Model for Predicting the testing set.
```{r p_data}
predict(rforest_model,test_data)
```

###Conclusion
Based on the analysis we found that there are many variables with empty values in the data set. These variables were omitted in the analysis also these variables could have an impact if the variable could have sufficient data in the dataset. Through analysis we also found that random forest model have better accuracy than decision tree model in the above analysis.